---
title: "ECN430 Assignment1"
author: "group 1"
date: 
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, eval=TRUE, echo=FALSE, warning=FALSE, error=FALSE}
rm(list = ls())
#install.packages("ggplot2")
#install.packages("zoo")
#install.packages("fpp2")
#install.packages("fUnitRoots")
#install.packages("vars")
library(dplyr)
library(ggplot2)
library(zoo)
library(tseries)
library(fpp2)
library(urca)
library(fUnitRoots)
library(vars)
library(forecast)
library(mFilter)
```



```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
# seasonally adjusted data
data <- read.csv("knr-tab-12-en.csv", header = F, sep = ";")
# select the useful data
gdp <- data[6,-(1:4)] %>% t() %>% na.omit(as.data.frame()) %>% as.numeric()
gdp <- ts(gdp, start = 1978, frequency = 4)

```


## a) Plot the following series and describe them

### GDP series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
# plot GDP series
autoplot(gdp/1000) + xlab("Time") + ylab("GDP:NOK billion") + 
  ggtitle("GDP series")

```

### The log transformed GDP series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
# log transformed GDP
lgdp <- ts(log(gdp), start = 1978, frequency = 4)

# plot log transformed GDP
autoplot(lgdp) + xlab("Time") + ylab("") + 
  ggtitle("Log transfromed GDP")

```

### The first difference of the log-series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
# first difference of the log-series
fdlog <- diff(lgdp, difference = 1)

# plot first difference of the log-series
autoplot(fdlog) + xlab("Time") + ylab("") + 
  ggtitle("First difference of the log-series")
ggsubseriesplot(fdlog) + xlab("Time") + ylab("First difference of the log-series")+ 
  ggtitle("Test of quarterly effect") 
```

### The YoY series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
# YoY growth rate
yoy <- diff(lgdp, lag = 4, differences = 1)

# plot the YoY (year-on-year %-age growth)
autoplot(yoy) + xlab("Time") + ylab("") + 
  ggtitle("YoY growth rate")
ggsubseriesplot(yoy) + ggtitle("Test of quarterly effect (YOY)") 

```

## b) Use a correlogram to test the stationary

### log-transformed GDP series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
ggAcf(lgdp, lag.max = 50)
ggPacf(lgdp)

```

### The first differenced log-transformed GDP series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
ggAcf(fdlog)
ggPacf(fdlog)

```

### YoY series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
ggAcf(yoy)
ggPacf(yoy)

```

## c) What about seasonal effects, do you have to take them away?

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
# unadjusted data
data2 <- read.csv("knr-tab-09-en.csv", header = F, sep = ";")
# select the useful data
ugdp <- data2[6,-(1:4)] %>% t() %>% na.omit(as.data.frame()) %>% as.numeric()
ugdp <- ts(ugdp, start = 1978, frequency = 4)

# compare seasonally adjusted and unadjusted data provided by SSB
autoplot(ugdp/1000, series = "unadjusted GDP series") + 
  autolayer(gdp/1000, series = "seasonally adjusted GDP series") + 
  xlab("Quarters") + ylab("billion NOK") + 
  ggtitle("Compare seasonally adjusted and unadjusted data provided by SSB")+ 
  theme(legend.position="bottom", legend.box = "vertical") 

# regress the unadjusted log GDP series on quarterly dummies and a trend variable
lugdp <- ts(log(ugdp), start = 1978, frequency = 4)
summary(tslm(lugdp ~ trend + season))
# save the residuals as a variable
reslugdp <- residuals(tslm(lugdp ~ trend + season))
# regress the seasonally adjusted GDP series on a trend variable
summary(tslm(lgdp ~ trend))
# save the residuals as a variable
reslgdp <- residuals(tslm(lgdp ~ trend))

# compare the residuals

autoplot(reslugdp, series = "Residuals of regressing
unadjusted log GDP 
from SSB on quarterly dummies 
and a trend variable") + 
  autolayer(reslgdp, series = "Residuals of regressing
seasonally adjusted log GDP 
from SSB on a trend variable") + 
  xlab("Quarters") + ylab("") + 
  ggtitle("Compare the residuals")+
  theme(legend.position = "bottom")

cor.test(reslugdp, reslgdp)

```


## d) Test whether the three series under question b) are stationary.

### Log transformed GDP series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.df(lgdp,type="trend", selectlags = "AIC"))
```


```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.kpss(lgdp,lags="long"))
```


```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.ers(lgdp, model = "trend"))
```

### The first differenced log-transformed GDP series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.df(fdlog,type="trend",selectlags = "AIC"))
```

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.kpss(fdlog,lags="long" ))
```

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.ers(fdlog, model = "trend"))
```

### YoY series

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.df(yoy,type="trend",selectlags = "AIC"))
```

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.kpss(yoy,lags="long"))
```

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
summary(ur.ers(yoy, model = "trend"))
```

## e) Explain what a partial autocorrelation function tells.

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
ggtsdisplay(yoy)
```


## f) Based on the information about the correlogram we try to fit an ARMA/ARIMA model.

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
Arima(yoy, order = c(2,0,1))
Arima(yoy, order = c(2,0,2))
Arima(yoy, order = c(2,0,3))
Arima(yoy, order = c(2,0,4))

```

## g) 
### Test the residual from your preferred model and comment.


```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
fit23 <- Arima(yoy, order = c(2,0,3))
checkresiduals(fit23)
coefficients(fit23)

fit24 <- Arima(yoy, order = c(2,0,4))
checkresiduals(fit24)
coefficients(fit24)

```

### Show the characteristic equation, and the characteristic roots from your preferred model.

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
coefficients(fit24)
cf <- c(0.5619339, 0.2711374, 1, 0)
cfmx <- matrix(data = cf, nrow = 2, byrow = T)
eigen(cfmx)
autoplot(fit24) + ggtitle("Inverse roots of characteristic polynomial")

```

## h) Compare the ARMA(2,2) model of yoy series to the one you preferred above. Briefly interpret your coefficients.


```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
Arima(yoy, order = c(2,0,4))
Arima(yoy, order = c(2,0,2))
```

## i) 
### Reestimate the ARMA(2,2) model and your preferred model until T=2007Q4.

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
#autoplot(forecast(fit))
yoy2 <- window(yoy, start=1978, end=c(2007,4))
Arima(yoy2, order = c(2,0,4))
Arima(yoy2, order = c(2,0,2))
```

###  Use the remaining observations until the end of the sample to generate a point forecast and a 95% confidence interval based on your estimated model. Briefly discuss your forecast.

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
yoy3 <- window(yoy, start=2008)

autoplot(forecast(Arima(yoy2, order =
  c(2,0,4)),h=length(yoy3)))+
  autolayer(yoy, series ="YOY-ARMA(2,4)" )+ 
  xlab("Time") + ylab("")+ 
  theme(legend.position="bottom", legend.box = "vertical")

autoplot(forecast(Arima(yoy2, order = c(2,0,2)),h=length(yoy3)))+
  autolayer(yoy, series ="YOY-ARMA(2,2)" )+ 
  xlab("Time") + ylab("")+ 
  theme(legend.position="bottom", legend.box = "vertical")
```

## j) 
### Evaluate the bias and RMSE for your preferred model, in the evaluation period (2008Q1 to 2018Q3).

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
accuracy(forecast(Arima(yoy2, order = c(2,0,4)),h=length(yoy3)),yoy3)

```

### Evaluate the bias and RMSE for the ARMA(2,2) model, in the evaluation period (2008Q1 to 2018Q3).

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
accuracy(forecast(Arima(yoy2, order = c(2,0,2)),h=length(yoy3)),yoy3)

```

### Again, what model do you prefer?

### And what do you conclude about the state of the Norwegian business cycle?

```{r, eval = TRUE, echo=FALSE, warning=FALSE, error=FALSE}
# use HP filter to deviate from trend estimate with HP filter
hplgdp <- hpfilter(lgdp, type = "lambda", freq = 40000)

autoplot(hplgdp$trend, series = "Trend") + 
  xlab("Time") + ylab("") + 
  ggtitle("HP filter") +
  autolayer(lgdp, series = "Actual")

autoplot(hplgdp$cycle) + xlab("Time") + ylab("") + 
  ggtitle("Cycle") + 
  geom_hline(yintercept = 0, colour = "black")

```